---
title: "sea lion 176 final"
format: pdf
pdf: default
gmf: default
editor: visual
---

# California Sea Lion Abundance Compared to Tide and Solar Radiation Data in 2025

**Authors**: Violet Lemley, Veronika Watkins, Piper Miller

## Question

How do tide height and solar radiation levels impact abundance of branded California Sea Lions on Ano Nuevo Island?

## Importance

Ano Nuevo Island is an important haul out zone for California Sea Lions (CSL). CSL can be at sea for weeks but heavily rely on haul outs to rest, molt, regulate body temperature, and most importantly, dry their fur. Dry fur is important to maintain its waterproofing, insulating coat and regulate body temperature, allowing oils in their skin to penetrate their fur. We wanted to see how tide height and average solar radiation impacted the number of branded CSL on the island.

This study is specifically looking at **branded** **CSL.** The observed individuals are branded by NOAA and are part of the West Coast Pinniped Branding program. They are captured, branded, released, and ideally never handled by humans again. Ideally, the branded CSL individuals would be a representative subset of how CSL would react to these factors.

```{r}
# read in tide and survery data
library(here)
library(brms)
library(hms)
library(tidyverse)
library(ggplot2) 
library(ggeffects)
library(dplyr)
library(lubridate)
tides <- read.csv("data/tides.csv")
surveys <- read.csv("data/surveys.csv")
```

## The Data

**NOAA Tide Height Data**

-   Ano Nuevo Island Station

-   05/01/2025-12/31/2025

**Average Solar Radiation Levels**

-   Ano Nuevo Dendra Meteorology Station

-   05/01/2025-08/31/2025

**Tagged California Sea Lion counts on Ano Nuevo Island**

-   05/01/2025- 08/25/2025

-   Monthly/Bi-monthly drone surveys of ANI

-   Visual analysis of surveys to identify branded individuals

-   Work done summer of 2025

## Data Set Editing

### Tide Data

Filtering out dates that surveys were not conducted.

```{r}
# converting the column to date format
tides$Date <- as.Date(tides$Date, format = "%Y/%m/%d")

# combine date and time into one column called date_time
tides$date_time <- ymd_hm(paste(tides$Date, tides$Time))

#creating vector with survey dates in order to filter out unecessary dates
target_dates <- as.Date(c("2025-05-01", "2025-05-12", "2025-05-28", "2025-06-05", "2025-06-17", "2025-06-30", "2025-07-02", "2025-07-07", "2025-07-15", "2025-07-24", "2025-07-29", "2025-08-12", "2025-08-15", "2025-08-25"))

#filtering out all other dates and saving as survey_tides
survey_tides <- tides %>%
  filter(Date %in% target_dates) %>%
  arrange(match(Date, target_dates))  
  
#view the results
print(survey_tides)
```

Combine Date and Time columns into DateTime to find specific tide height for each survey time and date.

```{r}

# convert Date and Time to correct format
Date = col_date(format = "")
Time = col_time(format = "")

# combine Date and Time columns into DateTime
survey_tides %>%
  mutate(
    # convert hms to period, then add to Date
    DateTime = as.POSIXct(Date) + period_to_seconds(as.period(Time))
  )

# remove Day, Date, and Time column
survey_tides <- survey_tides %>%
  select(-Day, -Date, -Time)

# view the results
head(survey_tides)
```

### CSL Count Data

Combining date and time into one column, DateTime. This is required to align with DateTime in tide data.

```{r}
#combining date and time into one column for CSL data 
surveys_CSL <- surveys %>%
 mutate(
   DateTime=parse_date_time(paste(Date,Time), "Ymd HMS") #formating time to 24 hr 
 )

#removing the seperate date and time columns 
surveys_CSL <- surveys_CSL %>%
  select(-Date, -Time)

# view results 
head(surveys_CSL)


```

### Solar Radiation Data

Filtering out dates and times that surveys were not conducted. Remaining data set will only include date and time of surveys.

```{r}
#reading in solar radiation data
solar <- read.csv("Solar Radiation.csv")

head(solar)

#filtering solar radiation data to survey dates and times using closest time available (recorded every 10 minutes)

#making vector with date and rounded times
target_dates_solar <- as.character(c("2025-05-01 17:40:00", "2025-05-12 09:30:00", "2025-05-28 10:40:00", "2025-06-05 08:30:00", "2025-06-17 08:20:00", "2025-06-30 08:50:00", "2025-07-02 10:10:00", "2025-07-07 09:40:00", "2025-07-15 09:30:00", "2025-07-24 08:30:00", "2025-07-29 09:00:00", "2025-08-12 09:20:00", "2025-08-15 08:50:00", "2025-08-25 10:00:00"))


#filtering entire data frame 
survey_solar <- solar %>%
  filter(Time %in% target_dates_solar) %>%
  arrange(match(Time, target_dates_solar)) 


```

### Combining Sea Lion Surveys and Tide Data

Pairing the CSL survey dates vector with their corresponding time stamps.

```{r}
#creating a vector of CSL survey times 
survey_time_vector <- c("17:43:30","09:32:42", "10:41:51", "08:28:14", "08:20:54", "08:52:25", "10:10:56", "09:38:41", "09:33:07", "08:27:45", "09:08:02", "09:18:32", "08:50:31", "09:59:10")

#created a vector where each CSL survey date is paired with its time stamp 
survey_times <- as.POSIXct(paste(target_dates, survey_time_vector),
                           format= "%Y-%m-%d %H:%M")


```

Using approx to find the approximate tide heights at each time listed in the vector.

```{r}
#doing a linear interpolation to get the approximate tide heights for the times of the CSL surveys, using the approx function

survey_tide_heights <- approx(
  x = as.numeric(tides$date_time),
                      #x is tide datetimes
  y = tides$Pred, 
                    #y is the tide heights
  xout= as.numeric(survey_times) 
  #xout is the CSL survey date times
) $y #extracts the interpolated tide heights from the output 

```

```{r}
#creaing a data frame of tide heights at each CSL survey time 
survey_table <- data.frame(
  Time = survey_times,  #column for times
  TideHeight = survey_tide_heights #column for heights corresponding to those heights 
)

#calling datafrane survey_table_complete
survey_table_complete <- data.frame(
  SurveyDateTime = survey_times, 
  TideHeight = survey_tide_heights, 
  CSLCount = surveys$Observed #number of CSLs observed 
)


#see the results
head(survey_table_complete)

```

### Adding in Solar Radiation

Separate date and time columns to be able to merge this data sheets with CSL+Tide without dealing with the rounded times.

```{r}
#survey_table_complete seperation
survey_table_complete.sep <- survey_table_complete %>%
# Separate the site_year column
separate(col = SurveyDateTime,
# Separate into two columns in this case
into = c("Date", "Time"),
# Separate by 11 characters in
sep = 11
)

#survey_solar seperation
survey_solar.sep <- survey_solar %>%
# Separate the site_year column
separate(col = Time,
# Separate into two columns in this case
into = c("Date", "Time"),
# Separate by 11 characters in
sep = 11
)
```

Merging solar radiation data and formatting final data set

```{r}
#combining both df by Date
merged_complete <- merge(survey_solar.sep, survey_table_complete.sep, by = "Date", all.x = TRUE)

#renaming solar radiation column
merged_complete <- merged_complete %>% rename("Solar Radiation" = "Año.Nuevo.Total.Solar.Radiation.Avg.W.m.2")

#getting rid of rounded time (Time.x) column
merged_complete <- merged_complete %>%
  select(-Time.x)

#rename Time.y to time
merged_complete <- merged_complete %>% rename("Time" = "Time.y")

#Reorder position of columns with time after date
merged_complete <- merged_complete %>% 
  relocate("Time", .before = "Solar Radiation")

head(merged_complete)
```

## Exploring the Data

### Tide Height

#### Scatter Plot

Scatter Plot of Tide Height Effect on CSL Count

```{r}
#tide height compared to observed CSL counts
tide.plot = ggplot(survey_table_complete, aes(x=TideHeight, y=CSLCount,)) +
  geom_point(color = "steelblue1") +
  geom_smooth(method=lm, color = "tan") +
  labs(title = "CSL Counts vs Tide Height", 
        x = "Tide Height (m)", 
        y = "CSL Count") + #adding a title and axis labels 
  theme_minimal()
print(tide.plot)

ggsave("tide.plot.png", plot = tide.plot, width = 10, height = 6, units = "in") # saving the png
```

We can see that their may be a slightly positive increase as CSL count does increase with tide height. This is just a initially visualization of the data. This Goes against our initial prediction that a lower tide height would show an increase in CSL abundance.

#### Comparing Tide Models

CSL Count as a Function of Tide Height- Trying a **Poisson Model**

```{r}
# CSL Count ~ Tide Height 
m.count.height.poi <- 
brm(data = survey_table_complete, #gave the model the data 
family = poisson, #choose a normal distribution
CSLCount ~ TideHeight,
iter = 2000, warmup = 1000, chains = 4, cores = 4,
file = "output/m.count.height/try")

#results 
summary(m.count.height.poi)
plot(m.count.height.poi)
```

```{r}
mean(survey_table_complete$CSLCount) # mean
var(survey_table_complete$CSLCount) # variance
```

This model ran correctly as the chains overlap each other and the posterior distributations are unimodal. Rhat is one but the 95%CI do overlap zero. Since the variance is much larger than the mean in this Poisson model, this means that this model is overdispersed.

Marine mammal researchers can sometimes use negative binomial models because marine mammal data frames are typically clustered and not random. This can lead to overdispersed count data in other model types. Negative binomial models account for that extra variability and can give a more reliable prediction.

Trying a **negative binomial model**:

```{r}
# CSL Count ~ Tide Height 
m.count.height.negbi <- 
brm(data = survey_table_complete, #gave the model the data 
family = negbinomial(), #choose a normal distribution
CSLCount ~ TideHeight,
iter = 2000, warmup = 1000, chains = 4, cores = 4,
file = "output/m.count.height/2nd")

#results 
summary(m.count.height.negbi)
plot(m.count.height.negbi)
```

This model showed that for each 1 meter increase in tide height multiplies the expected CSL count by about 1.73. The CI does contain zero, indicating substantial uncertainty. This could be due to our small sample size. We tried some stronger priors, but this still did not help because the data set is too small.

```{r}
loo(m.count.height.poi)
loo(m.count.height.negbi)
```

```{r}
waic(m.count.height.poi)
waic(m.count.height.negbi)
```

The negative binomial model is a better predictive fit for this data since the WAIC and LOOIC are lower compared to the Poisson model. This means this model will better predict our data than the Poisson model would.

#### Predictions

```{r}
preds.tides <- predict_response(m.count.height.negbi,
terms = c("TideHeight"))
plot(preds.tides, show_data = TRUE)
```

### Solar Radiation

#### Scatter Plot

```{r}
# rename Solar Radiation to Solar_Radiation
merged_complete <- merged_complete %>% rename("Solar_Radiation"= 'Solar Radiation')

# scatterplot of solar radiation effect on CSL count
radiation.plot = ggplot(merged_complete, aes(x=Solar_Radiation, y=CSLCount)) +
  geom_point(color = "steelblue1") +
  geom_smooth(method=lm, color = "tan") +
  labs( title = "CSL Count vs Solar Radiation", 
        x = "Solar Radiation (avg W/m^2)", 
        y = "CSL Count") #adding a title and axis label
print(radiation.plot)

ggsave("radiation.plot.png", plot = radiation.plot, width = 10, height = 6, units = "in") # saving the png
```

This plot shows that there might be a slight increase in CSL count as solar radiation increases. This is just our initial visualization. This agrees with our prediction that as solar radiation increases, so would CSL count.

#### Model

CSL Count as a Function of Solar Radiation

```{r}
# CSL Count ~ Solar Radiation 
m.count.solar.negbi <- 
brm(data = merged_complete, #gave the model the data 
family = negbinomial(), #negative binomial
CSLCount ~ Solar_Radiation,
iter = 2000, warmup = 1000, chains = 4, cores = 4,
file = "output/m.count.solar.negbi")

#results 
summary(m.count.solar.negbi)
plot(m.count.solar.negbi)
```

The model ran nice. This model shows no impact of solar radiation on CSL count. Large uncertainty due to small sample size though.

#### Predictions

```{r}
preds.int <- predict_response(m.CSL.Solar,
terms = c("Solar_Radiation"))
plot(preds.int, show_data = TRUE)
```

### Additive

#### Model

```{r}
# running additive model
m.CSL.Solar.combined <-
brm(data = merged_complete, #gave the model the data 
family = gaussian,#choose a normal distribution
CSLCount ~ Solar_Radiation + TideHeight,
iter = 2000, warmup = 1000, chains = 4, cores = 4,
file = "output/m.CSL.Solar.combined")

# summary stats
summary(m.CSL.Solar.combined)
plot(m.CSL.Solar.combined)

```

Again, shows no impact of solar radiation on CSL count, only tide. The estimate for solar radiation is 0.00

#### Predictions

```{r}
# predictions for additive model
preds.add <- predict_response(m.CSL.Solar.combined,
terms = c("TideHeight", "Solar_Radiation"))

plot(preds.add, show_data = TRUE)
```

```{r}
loo(m.CSL.Solar.combined)
loo(m.CSL.Solar)
```

PSIS analysis shows that the model only looking at solar radiation has better predictive power than the additive model. WAIC had a waic estimate greater than 0.4 so we used PSIS instead.

## Bias

Our study had a very limited sample size, so this it is difficult to use this to represent a larger population of CSL. Our surveys were also taken around the same time of day. All but one were taken between 8-11 am. Another potential bias could be the time of year. The surverys capture May through August which is a narrow view of the annual CSL cycle. Late June through early August is their breeding season.
