---
title: "sea lion 176 final"
format: pdf
pdf: default
gmf: default
editor: visual
---

```{r}
# read in tide and survery data
library(here)
library(brms)
library(hms)
library(tidyverse)
library(ggplot2) 
library(dplyr)
library(lubridate)
tides <- read.csv("data/tides.csv")
surveys <- read.csv("data/surveys.csv")
```

## Editing the Data sets

### Tides

We need to select tide data from the dates of surveys and get rid of all other dates, this new data set will be called survey_tides

```{r}
# converting the column to date format
tides$Date <- as.Date(tides$Date, format = "%Y/%m/%d")

# combine date and time into one column called date_time
tides$date_time <- ymd_hm(paste(tides$Date, tides$Time))

#creating vector with survey dates
target_dates <- as.Date(c("2025-05-01", "2025-05-12", "2025-05-28", "2025-06-05", "2025-06-17", "2025-06-30", "2025-07-02", "2025-07-07", "2025-07-15", "2025-07-24", "2025-07-29", "2025-08-12", "2025-08-15", "2025-08-25"))

#filtering out all other dates and saving as survey_tides
survey_tides <- tides %>%
  filter(Date %in% target_dates) %>%
  arrange(match(Date, target_dates))  
  
#view the results
print(survey_tides)
```

Combine Date and Time columns into DateTime to set up for finding specific tide height for each survey time and date

```{r}

# convert Date and Time to correct format
Date = col_date(format = "")
Time = col_time(format = "")

# combine Date and Time columns into DateTime
survey_tides %>%
  mutate(
    # convert hms to period, then add to Date
    DateTime = as.POSIXct(Date) + period_to_seconds(as.period(Time))
  )

# remove Day, Date, and Time column
survey_tides <- survey_tides %>%
  select(-Day, -Date, -Time)

# view the results
head(survey_tides)
```

<<<<<<< HEAD
![]()

### Seal Surveys

Combining date and time columns in seal survey data
=======
![]() Using aprox to find the approximate tide heights at each time listed in the vector
>>>>>>> 4417695 (not sure here)

```{r}
#combining date and time into one column for seal data 

surveys <- surveys %>%
 mutate(
   DateTime=parse_date_time(paste(Date,Time), "Ymd HMS p") #formating time to 24 hr 
 )

#removing the seperate date and time columns 
surveys <- surveys %>%
  select(-Date, -Time)

# view results 
head(surveys)


```

### Matching each seal survey to nearest tide measurement

#### Creating vectors

Pairing the seal survey dates vector with their corresponding time stamps

```{r}
#creating a vector of seal survey times 
survey_time_vector <- c("17:43:30","09:32:42", "10:41:51", "08:28:14", "08:20:54", "08:52:25", "10:10:56", "09:38:41", "09:33:07", "08:27:45", "09:08:02", "09:18:32", "08:50:31", "09:59:10")

#created a vector where each seal survey date is paired with its time stamp 
survey_times <- as.POSIXct(paste(target_dates, survey_time_vector),
                           format= "%Y-%m-%d %H:%M")


```

Using approx to find the approximate tide heights at each time listed in the vector

```{r}
#doing a linear interpolation to get the approximate tide heights for the times of the seal surveys, using the approx function

survey_tide_heights <- approx(
  x = as.numeric(tides$date_time),
                      #x is tide datetimes
  y = tides$Pred, 
                    #y is the tide heights
  xout= as.numeric(survey_times) 
  #xout is the seal survey date times
) $y #extracts the interpolated tide heights from the output 

#created a vector of tide heights at each seal survey time 
```

#### Creating a new table

```{r}
#creaing a data frame of tide heights at each seal survey time 
survey_table <- data.frame(
  SurveyTime = survey_times,  #column for times
  TideHeight = survey_tide_heights #column for heights corresponding to those heights 
)

# view results 
head(survey_table)
```

### Putting it all together 

```{r}
#creating a date frame with survey dates, survey times, tide heights, and seal count 
survey_table_complete <- data.frame(
  SurveyDateTime = survey_times, 
  TideHeight = survey_tide_heights, 
  SealCount = surveys$Observed #number of seals observed 
)
 #view the results 
head(survey_table_complete)
```

## Exploring the data

### Simple plots? 

```{r}
#tide height compared to observed seal counts
ggplot(survey_table_complete, aes(x=TideHeight, y=SealCount)) +
  geom_point() +
  geom_smooth(method=lm) +
  labs( title = "Seal Counts vs Tide Height", 
        x = "Tide Height", 
        y = "Seal Count") #adding a title and axis labels 
```

We can see that their may be a slightly positive increase as seal count does increase with tide height. This is just a initially visualization of the data.

#### Seal count as a function of tide height 

```{r}
# Seal Count ~ Tide Height 
m.count.height <-
brm(data = survey_table_complete, #gave the model the data 
family = gaussian,#choose a normal distribution
SealCount ~ 1 + TideHeight,
iter = 2000, warmup = 1000, chains = 4, cores = 4,
file = "output/m.count.height")

#results 
summary(m.count.height)
plot(m.count.height)
```
